---
title: "Nigeria Sample Report"
author: "Ruihan(Bonnie) Bao"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document
params:
  data: NA
  muilti: NA
  vactrend: NA
  dic_trend: NA,
  merged_data_for_vac_dis: NA
  preventable_vac_trend: NA
always_allow_html: true
---

# Comparison with other locations of similar geography or SDI

The construction of the Vaccine Improvement Index is based, in-part, on research on individual and socioeconomic factors associated with vaccine coverage in Nigeria. Prior research has found that mother’s age, education, and wealth as significantly related to immunization coverage after adjusting for other factors. In addition, the child’s birth order, family size, and place of delivery (home, public, or private facility) were related to vaccination coverage as well (1).


```{r, echo=FALSE}
library(knitr)
library(plotly)
library(dplyr)

multi = params$muilti
plotdata = data.frame(params$data)
plot = dplyr::filter(plotdata,plotdata$location %in% multi)

ggplot(data=plot, aes(year, result,group=location))+
   geom_line(aes(color=location))+
  geom_point(aes(color=location))+ 
  ggtitle("Time Series of Vaccine Improvement Index") + 
  xlab("Year") + ylab("VIX")
```

# Vaccination Trends

Between 2014 and 2019, Nigeria saw greater-than average improvements in seven routine vaccinations (out of 11 measured) (2). The progress demonstrated in this period contrasts to many years of stalled and even worsening vaccine coverage previously. Between 2005 and 2009, barriers to vaccination included structural issues including lack of security and armed conflict (3), supply chain and service delivery issues (4), and cultural and religious beliefs affecting vaccine hesitancy (5).

```{r, echo=FALSE}
plotdata = filter(params$vactrend,gsub(" ", "", location_name) == gsub(" ", "", "Nigeria"))
ggplot(data=plotdata, aes(year_id, prop_val,group=vaccine_name))+
   geom_line(aes(color=vaccine_name))+
  geom_point(aes(color=vaccine_name))+ 
  ggtitle("Time Series of Vaccination Coverage") + 
  xlab("Year") + ylab("Vaccination coverage (%)")
```

# Mortality and Disability Trends

Several vaccine-preventable diseases present a large burden on the population in Nigeria. For instance, by 2013 Nigeria was one of three countries in the world with endemic polio; yet, Nigeria also struggled with declining polio vaccine coverage (6).

```{r, echo=FALSE,warning=FALSE}
plotdata = filter(params$dic_trend,location_name == "Nigeria")

ggplot(data=plotdata, aes(year_id, ylds_number_val,group=cause_name))+
   geom_line(aes(color=cause_name))+
  geom_point(aes(color=cause_name))+ 
  ggtitle("Time Series of Years Lived in Less Than Ideal health in Population") + 
  xlab("Year") + ylab("Years lived with disability in population")+ 
  scale_y_continuous(trans='log2')+
  theme(legend.position="bottom",
        legend.text=ggplot2::element_text(size=7),
        legend.box.margin = margin(6, 6, 6, 6))+
  guides(fill=guide_legend(nrow=50))
```


# Relationship between Vaccines and Corresponding Diseases
Greater attention to polio and other vaccine-preventable diseases led to both improved vaccination coverage and decreases in the number of deaths from diseases like measles. Revised national strategic plans for polio and routine immunizations (2013-2015) have also allowed the country to implement additional evidence-based interventions and plans for routine immunization (3–5,7).

```{r, echo=FALSE,warning=FALSE}
selected_vac_data = filter(params$preventable_vac_trend, vaccine_trends$vaccine_name=="MCV1")
selected_vac_plotdata <- filter(selected_vac_data,location_name == "Nigeria")
dis_data_for_selected_vac = filter(params$merged_data_for_vac_dis, vaccine_name=="MCV1")
selected_dis_plotdata <- filter(dis_data_for_selected_vac,location_name == "Nigeria")
merged_selected_plotdata <- left_join(selected_vac_plotdata,selected_dis_plotdata, "year_id", "year_id")


# A few constants
deathColor <- "#69b3a2"
Color <- rgb(0.2, 0.6, 0.9, 1)

ggplot(merged_selected_plotdata, aes(x=year_id)) +
  geom_line(aes(y=round(deaths_rate_val,8)), size=1, color=deathColor) + 
  geom_line(aes(y=round(prop_val,2)/0.008), size=1, color=Color) +
  scale_y_continuous(
    # Features of the first axis
    name = "Measles:Deaths per 100,000 population",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans=~.*0.008, name="MCV 1 Vaccine coverage (%)")
  ) + 
  theme(
    axis.title.y = element_text(color = deathColor, size=13),
    axis.title.y.right = element_text(color = Color, size=13)
  ) +
  xlab("Year")+
  ggtitle("Vaccine & Corresponding Disease Trend")
```
