---
title: "Arizona Data Profile"
author: "UW PHI"
date: "04/01/2022"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# source set-up file
source("C:/Users/frc2/Documents/uw-phi-vax/resilient_imm_sys/aim_1/03_data_profiles/01_set_up_R.r")

s <- "Arizona" # state
c <- c("Navajo County", "Pinal County", "Gila County") # counties
r <- "arizona" # region
sr <- c("navajo", "pinal", "gila") # subregion
```

# 1. State-level trends

```{r}
# load state level data
state_data <- readRDS(paste0(prepped_data_dir, "11_merged_data_for_state_profile_docs.RDS")) %>%
  # filter state
  filter(state==s)

# reshape data for plotting trends in vaccination gaps
state_data_gaps <- state_data %>% 
  select(state, YEAR, income, VACCINE, hispa_diff, black_diff, other_diff) %>%
  pivot_longer(cols = c(hispa_diff, black_diff, other_diff),
               names_to = "variable",
               values_to = "value") %>%
  mutate(variable_label = case_when(
    variable=="hispa_diff" ~ "Hispanic",
    variable=="black_diff" ~ "Black/African-American",
    variable=="other_diff" ~ "Other or multi-racial"
  )) %>% 
  mutate(income_label = case_when(
    income=="high" ~ 1,
    income=="med" ~ 2,
    income=="low" ~ 3 ,
    income=="miss" ~ 4,
  ))

# factor income_label
state_data_gaps$income_label <- factor(state_data_gaps$income_label,
                                       labels = c("High income", 
                                                  "Medium income", 
                                                  "Low income", 
                                                  "Missing income "))
  

# plot trends in vaccination coverage among different racial/ethnic groups
ggplot(state_data_gaps, aes(YEAR, value, group=factor(VACCINE))) + 
  geom_line(aes(color=VACCINE), show.legend = TRUE) +
  facet_grid(vars(income_label), vars(variable_label), labeller = label_value) + 
   labs(title=paste('Differences in percent of children fully vaccinated in each racial/ethnic group'), 
       subtitle=paste0('compared to white children of same income background in', " ", s),
       caption="Data Source: 2007 and 2019 National Immunization Survey (NIS)") +
  ylab('Percentage points') + 
  scale_x_continuous(breaks=c(2007, 2019)) +
  xlab('Year') +
  ylim(-.06, .06)
  # theme_minimal()

```

We calculated percent coverage using data from the National Immunization Survey which provides information on how many children in the state received all recommended vaccines. For each year we compared the vaccination coverage among children who were Black, Hispanic, or Other/multiple-race background to White children of the same income level. A gap greater than 0 suggests that white children are being vaccinated at a higher rate than the group of non-white children they are compared with. In the graphs above, a positive slope indicates increasing differences (worse gap), and a negative slope indicates decreasing differences (an improvement in the gap). 

*What this graphic shows is that there is trend of an increasing gap in vaccination coverage for Black children compared to White children for the DTP vaccine. For the MMR Vaccine, White children had lower vaccination coverage rates in 2007 than Hispanic children, but the gap has decreased in 2019. For children that are in the "other" or "multi-racial" category the gap is growing for all three vaccines however. Unfortunately, lack of further race/ethnicity disaggregation makes it difficult to identify the background of the children in these categories.*

# 2. State-level trends by race/ethnicity and income

```{r}
state_cov <- state_data %>% select(state, YEAR, income, VACCINE, white, hispa, black, other) %>% 
  pivot_longer(cols = c(white, hispa, black, other),                                                                                          
               names_to = "race",                                                                                           
               values_to = "value") %>%
    mutate(race_label = case_when(
    race=="hispa" ~ "Hispanic",
    race=="black" ~ "Black",
    race=="other" ~ "Other/multiple",
    race=="white" ~ "White"
  ))

state_cov$value <- as.numeric(state_cov$value)

# factor the income level
state_cov <- state_cov %>% 
  mutate(income_label = case_when(
    income=="high" ~ 1,
    income=="med" ~ 2,
    income=="low" ~ 3,
    income=="miss" ~ 4
  ))

# factor income_label
state_cov$income_label <- factor(state_cov$income_label,
                                       labels = c("High", 
                                                  "Medium", 
                                                  "Low", 
                                                  "Missing"))

# create raster plot of locations
ggplot(state_cov, aes(income_label, VACCINE)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=median, size=1, fill=NA, colour="black") +
  labs(title=paste('Percent of children fully vaccinated'), y = 'Vaccine', x = 'Income level', 
       subtitle=paste0('in', " ", s),
       caption="Data Source: 2007 and 2019 National Immunization Survey (NIS)", fill='Darker value indicates \n higher vaccination rate') +
  facet_grid(vars(race_label), vars(YEAR)) +
  theme_minimal()
  
```

This graphic provides the actual immunization rates as estimated using the NIS Survey and used in calculating the gaps according to race and ethnicity from the first graphic. 

*What this graphic shows is that for some vaccines such as DTP and Hep B, coverage levels dropped in 2019. The pattern is similar across racial/ethnic and income groups. However, Black children had slightly larger drops in coverage than White children for the DTP vaccine in particular, resulting in an increased vaccination gap pointed out in Figure 1. Similarly, although vaccination coverage improved among White children for the MMR vaccine, among children in the "Other/multi-race" group, MMR coverage actually stayed the same or slightly decreased.*

# 3. Map of county locations and populations

```{r}
# load data to be used
data <- readRDS(paste0(prepped_data_dir, "12_merged_data_for_county_profile_docs.RDS"))

# subset state and counties of interest
data <- data %>% filter(state==s & county %in% c)

# select columns of interest and variables to plot
plot_data <- data %>% filter(variable=="total_population") %>%
  select(county, fips, value)

plot_data$value <- as.numeric(plot_data$value)

# create dataframe of state level mapping shapefile data
all_state_df <- map_data("state")
state_df <- subset(all_state_df, region==r)

# create dataframe of county level mapping shapefile data
all_county_df <- map_data("county")
subset_county <- subset(all_county_df, region==r)

plot_data$subregion <- tolower(plot_data$county)
subset_county$subregion <- paste0(subset_county$subregion, " county")

plot_data <- left_join(plot_data, subset_county, by=c("subregion"))
plot_data$subregion <- str_to_title(plot_data$subregion) 

ggplot(data=state_df, mapping=aes(x=long, y=lat, group=group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color="black", fill="gray") +
  geom_polygon(data=subset_county, fill=NA, color="white") +
  geom_polygon(data = plot_data, mapping = aes(x=long, y=lat, group=group, fill=value))+
  scale_fill_continuous(name = "Population (2019)", label = scales::comma) +
  geom_text(data= plot_data[c(46, 158, 213),], label=plot_data[c(46, 158, 213),]$subregion, color="white", size=4, fontface="bold") +
  labs(title=paste0('Population of counties selected in ', s),
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates") +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())
```

*Counties we selected are all in the central part of the state.*

# 4. Demographics of children under 4 in each county

```{r}

demo_data <- data %>% filter(variable %in% c("Non-Hispanic White",
 "African-American/Black",
 "American Indian",
 "Asian-American",
  "Hispanic"))

demo_data$value <- as.numeric(demo_data$value)

ggplot(demo_data, aes(fill=variable, y=value, x=county)) + 
  geom_bar(position = "fill", stat="identity") + 
  scale_fill_brewer(palette = "Paired") + 
  labs(title=paste('Demographics of children under 4 in', s, 'in 2019'), y = 'Relative proportions', x = 'County',
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates", fill="Race/ethnicity") +
  theme_minimal()
```

# Childhood Vaccination data

```{r}
county_data <- data %>% filter(variable %in% c("Fully vaccinated children in 2015",
                                         "Fully vaccinated children in 2019",
                                         "Change between time points")) %>% 
  select(county, variable, value) %>% 
  mutate(variable = recode(variable, 
                           "Fully vaccinated children in 2015"="2015",
                           "Fully vaccinated children in 2019"="2019",
                            "Change between time points"="change"))

county_data$value <- as.numeric(county_data$value)

# create raster plot of locations
ggplot(county_data, aes(variable, county)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Vaccination coverage among children'), y = 'County', x = 'Year', 
       subtitle=paste0('comparing 2015 and 2019, in', " ", s),
       caption="Data Source: State Department of Health", fill='Darker value indicates \n higher vaccination rate') +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()
```

Negative "change" values indicate that the proportion of children vaccinated decreased between the two time periods. 

*Navajo County's improvement of 2 percentage points is high compared to the average change seen in counties across the state.*

# 5. Covid vaccination among adults

```{r}

county_covid_data <- readRDS(paste0(prepped_data_dir, "10_prepped_county_vaccination_data_disaggregated.RDS")) %>% 
  filter(state == s) %>% 
  filter(county %in% c)

#  re-code the variables of race/ethnicity so they fit on the charge
county_covid_data <- county_covid_data %>% mutate(race_label=case_when(
  race_ethnicity=="Unknown" ~ "Not Reported",
  race_ethnicity=="Other" ~ "Not Reported",
  race_ethnicity=="White" ~ "Non-Hispanic White",
  race_ethnicity=="Black or African-American" ~ "African-American/Black",
  race_ethnicity=="Asian or Pacific Islander" ~ "Asian-American",
  TRUE ~ race_ethnicity
))

if (s=="Arizona"){
  ggplot(county_covid_data, aes(fill=race_label, y=proportion, x=county)) + 
  geom_bar(position = "fill", stat="identity") + 
  scale_fill_brewer(palette = "Paired") + 
  labs(title=paste('Race/ethnicity among those that received first Covid dose', s), y = 'Relative proportions', x = 'County',
       subtitle="as of March 19, 2022",
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates", fill="Race/ethnicity") +
  theme_minimal()
  
} else {
  ggplot(county_covid_data, aes(race_ethnicity, county)) +
  geom_tile(aes(fill = proportion)) + 
  geom_text(aes(label = round(proportion, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Proportion of adults vaccinated'), y = 'County', x = 'Race/ethnicity', 
       subtitle=paste0('by race and ethnicity, in', " ", s),
       caption="Data Source: State Department of Health \n retrieved week of March 21, 2022", fill='Darker value indicates \n higher vaccination rate') +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()
}

```

*In Arizona, county vaccination levels are not available by race/ethnic group. What this graphic shows instead is how the first vaccination dose has been distributed so far. At the county level, vaccination distribution mostly aligns with population distribution. The only exception seems to be Gila county. Although Native American's make up the majority of the adult population in Gila County (20% of the population is Native American), only about 13% of the first doses went to Native Americans. *

# 6. County rankings

```{r}
ranking_data <- data %>% filter(variable %in% c("County level of childhood vaccination", "County level of Covid-19 vaccination")) %>% 
  mutate(variable_label = case_when(
    variable=="County level of childhood vaccination" ~ "childhood vaccination",
    variable=="County level of Covid-19 vaccination" ~ "Covid-19 vaccination"
  ))

ranking_data$value <- factor(ranking_data$value,
                             levels = c("low", "medium", "high"))
ggplot(ranking_data, aes(variable_label, county)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = value)) +
  # scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Between county-comparisons on childhood and covid vaccines '), y = 'County', x = '', 
       subtitle=paste0('Rankings based on vaccination levels across all counties in', " ", s),
       caption="Data Source: State Health Department and CDC Covid Data Tracker \n retrieved March 18, 2022",
       fill="level of vaccination") +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()

```

County grouping was determined by ranking all counties according to the proportion of population vaccinated and dividing them into three groups: those in the top 25% were classified as high performers, those in the bottom 25% were classified as low performers, and the remaining were medium performers. The same process was used for both routine childhood immunization data and Covid-19 vaccination data. 

*What this graphic shows is that level of childhood vaccination coincided with levels of Covid-19 vaccination in at least two cases: Navajo County was in the "high" group for both childhood and Covid-19 vaccination rates and Pinal County was in the "medium" grouping for both. *
