---
title: "Washington Data Profile"
author: "UW PHI"
date: "04/01/2022"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# source set-up file
source("C:/Users/frc2/Documents/uw-phi-vax/resilient_imm_sys/aim_1/03_data_profiles/01_set_up_R.r")

s <- "Washington" # state
c <- c("Skagit County", "Yakima County", "Chelan County") # counties
r <- "washington" # region
sr <- c("skagit", "yakima", "chelan") # subregion
```

# 1. State-level trends

```{r}
# load state level data
state_data <- readRDS(paste0(prepped_data_dir, "11_merged_data_for_state_profile_docs.RDS")) %>%
  # filter state
  filter(state==s)

# reshape data for plotting trends in vaccination gaps
state_data_gaps <- state_data %>% 
  select(state, YEAR, income, VACCINE, hispa_diff, black_diff, other_diff) %>%
  pivot_longer(cols = c(hispa_diff, black_diff, other_diff),
               names_to = "variable",
               values_to = "value") %>%
  mutate(variable_label = case_when(
    variable=="hispa_diff" ~ "Hispanic",
    variable=="black_diff" ~ "Black/African-American",
    variable=="other_diff" ~ "Other or multi-racial"
  )) %>% 
  mutate(income_label = case_when(
    income=="high" ~ 1,
    income=="med" ~ 2,
    income=="low" ~ 3 ,
    income=="miss" ~ 4,
  ))

# factor income_label
state_data_gaps$income_label <- factor(state_data_gaps$income_label,
                                       labels = c("High income", 
                                                  "Medium income", 
                                                  "Low income", 
                                                  "Missing income "))
  

# plot trends in vaccination coverage among different racial/ethnic groups
ggplot(state_data_gaps, aes(YEAR, value, group=factor(VACCINE))) + 
  geom_line(aes(color=VACCINE), show.legend = TRUE) +
  facet_grid(vars(income_label), vars(variable_label), labeller = label_value) + 
   labs(title=paste('Differences in percent of children fully vaccinated in each racial/ethnic group'), 
       subtitle=paste0('compared to white children of same income background in', " ", s),
       caption="Data Source: 2007 and 2019 National Immunization Survey (NIS)") +
  ylab('Percentage points') + 
  scale_x_continuous(breaks=c(2007, 2019)) +
  xlab('Year') +
  ylim(-.06, .06)
  # theme_minimal()

```

We calculated percent coverage using data from the National Immunization Survey which provides information on how many children in the state received all recommended vaccines. For each year we compared the vaccination coverage among children who were Black, Hispanic, or Other/multiple-race background to White children of the same income level. A gap greater than 0 suggests that white children are being vaccinated at a higher rate than the group of non-white children they are compared with. In the graphs above, a positive slope indicates increasing differences (worse gap), and a negative slope indicates decreasing differences (an improvement in the gap). 

*African-American children are vaccinated at lower rates than White children in Washington state and that seems to have changed very little during the two time points for DTP and HEP B vaccines. However, for MMR the vaccination gap seems to be increasing. In 2007, Hispanic children seemed to have higher vaccination coverage than White children, although that gap has narrowed in recent years and both groups have similar coverage in the most recent survey. Children from other racial backgrounds have a slight gap in vaccination coverage, but it's not as large as the gap between African-American and White children. *

# 2. State-level trends by race/ethnicity and income

```{r}
state_cov <- state_data %>% select(state, YEAR, income, VACCINE, white, hispa, black, other) %>% 
  pivot_longer(cols = c(white, hispa, black, other),                                                                                          
               names_to = "race",                                                                                           
               values_to = "value") %>%
    mutate(race_label = case_when(
    race=="hispa" ~ "Hispanic",
    race=="black" ~ "Black",
    race=="other" ~ "Other/multiple",
    race=="white" ~ "White"
  ))

state_cov$value <- as.numeric(state_cov$value)

# factor the income level
state_cov <- state_cov %>% 
  mutate(income_label = case_when(
    income=="high" ~ 1,
    income=="med" ~ 2,
    income=="low" ~ 3,
    income=="miss" ~ 4
  ))

# factor income_label
state_cov$income_label <- factor(state_cov$income_label,
                                       labels = c("High", 
                                                  "Medium", 
                                                  "Low", 
                                                  "Missing"))

# create raster plot of locations
ggplot(state_cov, aes(income_label, VACCINE)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=median, size=1, fill=NA, colour="black") +
  labs(title=paste('Percent of children fully vaccinated'), y = 'Vaccine', x = 'Income level', 
       subtitle=paste0('in', " ", s),
       caption="Data Source: 2007 and 2019 National Immunization Survey (NIS)", fill='Darker value indicates \n higher vaccination rate') +
  facet_grid(vars(race_label), vars(YEAR)) +
  theme_minimal()
  
```

This graphic provides the actual immunization rates as estimated using the NIS Survey and used in calculating the gaps according to race and ethnicity from the first graphic. 

*What this graphic shows is that DTP vaccines consistently have the lowest rates of coverage among children from all races or ethnicity compared to other vaccines in the 2007 and 2019 surveys. In the most recent survey, both Black and White children from a high-income background have seen improved vaccination coverage for the HEP B and DTP vaccines.*

# 3. Map of county locations and populations

```{r}
# load data to be used
data <- readRDS(paste0(prepped_data_dir, "12_merged_data_for_county_profile_docs.RDS"))

# subset state and counties of interest
data <- data %>% filter(state==s & county %in% c)

# select columns of interest and variables to plot
plot_data <- data %>% filter(variable=="total_population") %>%
  select(county, fips, value)

plot_data$value <- as.numeric(plot_data$value)

# create dataframe of state level mapping shapefile data
all_state_df <- map_data("state")
state_df <- subset(all_state_df, region==r)

# create dataframe of county level mapping shapefile data
all_county_df <- map_data("county")
subset_county <- subset(all_county_df, region==r)

plot_data$subregion <- tolower(plot_data$county)
subset_county$subregion <- paste0(subset_county$subregion, " county")

plot_data <- left_join(plot_data, subset_county, by=c("subregion"))
plot_data$subregion <- str_to_title(plot_data$subregion) 

ggplot(data=state_df, mapping=aes(x=long, y=lat, group=group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color="black", fill="gray") +
  geom_polygon(data=subset_county, fill=NA, color="white") +
  geom_polygon(data = plot_data, mapping = aes(x=long, y=lat, group=group, fill=value))+
  scale_fill_continuous(name = "Population (2019)", label = scales::comma) +
  geom_text(data= plot_data[c(43, 106, 194),], label=plot_data[c(43, 106, 194),]$subregion, color="white", size=4, fontface="bold") +
  labs(title=paste0('Population of counties selected in ', s),
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates") +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())
```

*Counties we selected are all in the north or eastern part of the state.*

# 4. Demographics of children under 4 in each county

```{r}

demo_data <- data %>% filter(variable %in% c("Non-Hispanic White",
 "African-American/Black",
 "American Indian",
 "Asian-American",
  "Hispanic"))

demo_data$value <- as.numeric(demo_data$value)

ggplot(demo_data, aes(fill=variable, y=value, x=county)) + 
  geom_bar(position = "fill", stat="identity") + 
  scale_fill_brewer(palette = "Paired") + 
  labs(title=paste('Demographics of children under 4 in', s, 'in 2019'), y = 'Relative proportions', x = 'County',
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates", fill="Race/ethnicity") +
  theme_minimal()
```

# 5. Childhood Vaccination data

```{r}
county_data <- data %>% filter(variable %in% c("Fully vaccinated children in 2015",
                                         "Fully vaccinated children in 2017",
                                         "Change between time points")) %>% 
  select(county, variable, value) %>% 
  mutate(variable = recode(variable, 
                           "Fully vaccinated children in 2015"="2015",
                           "Fully vaccinated children in 2017"="2017",
                            "Change between time points"="change"))

county_data$value <- as.numeric(county_data$value)

# create raster plot of locations
ggplot(county_data, aes(variable, county)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Vaccination coverage among children'), y = 'County', x = 'Year', 
       subtitle=paste0('comparing 2015 and 2017, in', " ", s),
       caption="Data Source: State Department of Health", fill='Darker value indicates \n higher vaccination rate') +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()
```
```{r}
# create another visual that is helpful for Washington state since those years are not comparable
county_data_two <- data %>% filter(variable %in% c("Fully vaccinated children in 2018",
                                         "Fully vaccinated children in 2019")) %>% 
  select(county, variable, value)

county_data$value <- as.numeric(county_data$value)

# reshape data
county_data_two <- pivot_wider(county_data_two, id_cols = county, names_from = variable, values_from = value)
county_data_two$`Fully vaccinated children in 2019` <- as.numeric(county_data_two$`Fully vaccinated children in 2019`)
county_data_two$`Fully vaccinated children in 2018` <- as.numeric(county_data_two$`Fully vaccinated children in 2018`)

county_data_two$change <- county_data_two$`Fully vaccinated children in 2019`-county_data_two$`Fully vaccinated children in 2018`

# reshape data
county_data_two <- pivot_longer(county_data_two, !county, names_to = "variable" )

county_data_two <- county_data_two %>% mutate(variable = recode(variable, 
                           "Fully vaccinated children in 2018"="2018",
                           "Fully vaccinated children in 2019"="2019",
                            "Change between time points"="change"))



# create raster plot of locations
ggplot(county_data_two, aes(variable, county)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Vaccination coverage among children'), y = 'County', x = 'Year', 
       subtitle=paste0('comparing 2018 and 2019, in', " ", s),
       caption="Data Source: State Department of Health", fill='Darker value indicates \n higher vaccination rate') +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()
```

Negative "change" values indicate that the proportion of children vaccinated decreased between the two time periods. Technical Note: Numbers for 2018 are not comparable to past years due to a change in how population data was measured.

*Skagit County saw an improvement of 15 percentage points between 2015 and 2017, a high value compared to other counties in the state. But a decrease of about 5% between 2018 and 2019.*

# 6. Covid vaccination among adults

```{r}

county_covid_data <- readRDS(paste0(prepped_data_dir, "10_prepped_county_covid_vaccination_data_disaggregated.RDS")) %>% 
  filter(state == s) %>% 
  filter(county %in% c)

#  re-code the variables of race/ethnicity so they fit on the charge
county_covid_data <- county_covid_data %>% mutate(race_label=case_when(
  race_ethnicity=="Black or African-American" ~ "Black",
  race_ethnicity=="Native Hawaiian/Pacific Islander" ~ "NHPI",
  race_ethnicity=="American Indian" ~ "AIAN",
  TRUE ~ race_ethnicity
))

if (s=="Arizona"){
  ggplot(county_covid_data, aes(fill=race_label, y=proportion, x=county)) + 
  geom_bar(position = "fill", stat="identity") + 
  scale_fill_brewer(palette = "Paired") + 
  labs(title=paste('Race/ethnicity among those that received first Covid dose', s), y = 'Relative proportions', x = 'County',
       subtitle="as of March 19, 2022",
       caption="Data Source: US Census Bureau \n 2021 County and Resident Population Estimates", fill="Race/ethnicity") +
  theme_minimal()
  
} else {
  ggplot(county_covid_data, aes(race_label, county)) +
  geom_tile(aes(fill = proportion)) + 
  geom_text(aes(label = round(proportion, 2))) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Proportion of adults vaccinated'), y = 'County', x = 'Race/ethnicity', 
       subtitle=paste0('by race and ethnicity, in', " ", s),
       caption="Data Source: State Department of Health \n retrieved week of March 21, 2022 \n AIAN: American Indian or Alaska Native \n NHPI: Native Hawaiian and Pacific Islander", fill='Darker value indicates \n higher vaccination rate') +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()
}
```

*What this graphic shows is that at the county level, there existed some large differences in Covid vaccination rates between racial/ethnic groups as well as between counties. In Yakima, Asian, Black, and White residents had the lowest vaccination rates (around 50%). While in Skagit county, Hispanic residents had the lowest vaccination rates. In Chelan county, most racial/ethnic groups has similar vaccination rates, except for American Indian residents which had lower rates (between 9 and 14 percentage points lower than other racial/ethnic groups).*

# 7. County rankings

```{r}
ranking_data <- data %>% filter(variable %in% c("County level of childhood vaccination", "County level of Covid-19 vaccination")) %>% 
  mutate(variable_label = case_when(
    variable=="County level of childhood vaccination" ~ "childhood vaccination",
    variable=="County level of Covid-19 vaccination" ~ "Covid-19 vaccination"
  ))

ranking_data$value <- factor(ranking_data$value,
                             levels = c("low", "medium", "high"))
ggplot(ranking_data, aes(variable_label, county)) +
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = value)) +
  # scale_fill_gradient(low = "white", high = "steelblue") +
  # geom_tile(data=county_data  %>% filter(variable=="change"), size=1, fill=NA, colour="black") +
  labs(title=paste('Between county-comparisons on childhood and covid vaccines '), y = 'County', x = '', 
       subtitle=paste0('Rankings based on vaccination levels across all counties in', " ", s),
       caption="Data Source: State Health Department and CDC Covid Data Tracker \n retrieved March 18, 2022",
       fill="level of vaccination") +
  # facet_grid(vars(VACCINE), vars(YEAR)) +
  theme_minimal()

```

County grouping was determined by ranking all counties according to the proportion of population vaccinated and dividing them into three groups: those in the top 25% were classified as high performers, those in the bottom 25% were classified as low performers, and the remaining were medium performers. The same process was used for both routine childhood immunization data and Covid-19 vaccination data. 

*What this graphic shows is that level of childhood vaccination does not always coincide with levels of vaccination for Covid-19. Yakima County was the exception as it seemed to fall within the "medium" group for both. Chelan county was in the "low" group for childhood vaccination but considered a county with high vaccination coverage for Covid-19 vaccines.*
